  ---
title: "A Study on the Factors that Affect the Average Serve Speed of Top Tennis Players"
author: "GJornert, EMelindo, HArnold"
date: "2026-02-02"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(wooldridge)
library(broom)
library(texreg)
library(readxl)
library(xtable)
library(dplyr)
library(corrr)

library(lmtest)
library(sandwich)
library(hettest)
library(vars)

library(nlme)

knitr::opts_chunk$set(echo = TRUE)

```

### The Determinants of Serve Speed in Professional Tennis 

In tennis, the serve is far more than just the starting point of a rally — it is one of the most decisive shots in the game. A powerful serve can earn a player a free point, limit their opponent's ability to return aggressively, and dictate the flow of the entire match. At the highest levels of professional tennis, serve speed is considered a critical competitive advantage, which raises an important question: what factors actually determine how fast an ATP player can serve?

This paper investigates the determinants of average serve speed among professional ATP players through a series of regression analyses. We examine both physical characteristics — including height, weight, and age — as well as experiential and equipment-related variables, such as years on tour and racket specifications. Our central research question is: to what extent do a player's physical attributes, professional experience, and equipment explain variation in average serve speed? More specifically, we ask whether natural physical factors or acquired, controllable factors are the stronger predictors of performance.

Understanding what drives serve speed also carries meaningful economic implications. If serve speed is primarily determined by physical traits, it becomes a more predictable variable for sports analysts, bettors, and scouts — offering an edge in forecasting match outcomes based on player profiles. On the other hand, if equipment plays a significant role, racket manufacturers have a tangible, data-backed basis for marketing performance claims, much as footwear and apparel brands have done in other sports. Either finding has real consequences for how money flows through the tennis industry.

We hypothesize that physical attributes — particularly height — will be the strongest predictor of serve speed, as taller players benefit from a more favorable serving angle and a longer swing arc, allowing them to generate greater velocity more naturally. Weight and age are also expected to be significant physical factors, with serve speed likely peaking during a player's prime years and declining thereafter. While we expect racket specifications — such as racket weight and string tension — to have a meaningful impact on serve speed as well, we anticipate their effect to be secondary to that of physical traits, since equipment can optimize performance but cannot fully compensate for the bio mechanical advantages conferred by a player's body. To test these hypotheses, our data set includes player-level statistics from the ATP tour, comprising physical measurements, equipment specifications, and career statistics.
Our findings confirm that height is the dominant and most robust determinant of serve speed, while most other variables — including age, equipment, and handedness — do not reach statistical significance in the full sample.

### Literature Review

We were not surprised to find out that other researches had previously been interested in trying to understand how different variables can potentially affect a player's serve speed.

#### Tennis Serve Biomechanics

Tennis Serve Biomechanics, Joint Load Mechanics, and Overuse Injuries: A Narrative Review by Caroline Martin, Pierre Touzard, Loic Fourel, and Babette M. Pluim examines how the body is affected by the act of serving. finding that improper form and technique induce eminently traumatic mechanical joint loading, which can lead to injury over time. To mitigate joint damage, the paper finds that teaching proper serving technique, choosing appropriate equipment, and training conditions can reduce the effects of a tennis serve on the body. providing a reason behind the link between a player's serve, their equipment, and their body.

#### The Role of Toss Zenith and Impact Height in the Serve Speed in Junior Tennis Players.

The Role of Toss Zenith and Impact Height in the Relationship Between Shoulder Rotation Strength and Serve Speed in Junior Tennis Players, by Jan Vacek, Michal Vagner, Jan Malecek, Jachym Simsa, Vladan Olah, and Petr Stastny. The paper details how various factors, such as zenith height, impact height, and the zenith–impact differential of an opening affect Serve Speed in Junior Tennis Players. Since this research is based on junior players that are transitioning, or trying to, into the professional world, the findings are likely to be similar to those of professional players. Their findings show that the height of the Impact showed a positive association with serve speed, as well as with the execution of the opening toss and its consistency, significantly increasing serve speed.

#### Age and sex-related upper body performance in young tennis players.

Age- and sex-related upper-body performance differences in competitive young tennis players, by Jaime Fernandez-Fernandez, Fabio Yuzo Nakamura, Victor Moreno-Perez, Alejandro Lopez-Valenciano, Juan Del Coso, and Cesar Gallo-Salazar. The study aimed to analyze the physical abilities of young male and female tennis players to identify relationships among physical variables and serve speed. The study directly uses serve speed as a factor, making it a relevant reference for comparing with our own findings. The study found that service speed increased across all players, regardless of sex.

### Data and variables

```{r} 
# Partial data
tennis.data <- read_excel("econ-research_test.xlsx")
```
```{r}
#rename column by name
colnames(tennis.data)[colnames(tennis.data) == '1serve'] <- 'frtserve'
colnames(tennis.data)[colnames(tennis.data) == '2fault'] <- 'dblfault'

```

```{r}
# Completed data set
fulldata <- read_excel("econ.research_v1.xlsx")
```
```{r}
#rename column by name
colnames(fulldata)[colnames(fulldata) == '1serve'] <- 'frtserve'
colnames(fulldata)[colnames(fulldata) == '2faults'] <- 'dblfault'

```

The data set used in this study draws from multiple sources, including the official ATP Tour website, individual tournament websites, and supplementary research databases. All observations reflect statistics recorded over the 2025 ATP season. The sample consists of the top 100 ranked ATP players as of year-end 2025, providing a comprehensive cross-section of professional men's tennis at its highest competitive level.

The average player in our sample is 26.7 years old and stands 187.4 cm tall (approximately 6'2") consistent with the physical profile of a modern ATP professional. Average serve speed across the sample is 189.1 km/h, with a minimum of 171 km/h and a maximum of 214 km/h, illustrating meaningful variation in serving power even among the world's best players. Players in the sample played an average of 43 matches during the season, with racket head sizes clustering between 600 and 645 cm², as the equipment is getting increasingly more standardized at the professional level. The vast majority of players in the sample are right-handed, which we believe will limit the statistical significance of handedness as a predictor. 

Table 1 below summarizes the key descriptive statistics for the full sample.

```{r}
summary(fulldata)
```

Two variables initially considered for inclusion were ultimately dropped prior to estimation. Years on tour was excluded due to its strong collinearity with age: players who have been on tour longer are, by construction, older, and including both variables would introduce multicollinearity that affects standard errors and complicates the interpretation. Gender was initially considered to allow for a broader sample, including WTA players, who systematically serve at lower speeds than their ATP counterparts. However, it was dropped due to data availability constraints on the WTA side, and the analysis was restricted to ATP players only. The final dataset includes 100 observations across 9 variables.

### Variable Descriptions

* $avgspeed$: Average serve speed (measured in **km/h**). This is the dependent variable, measured as each player's average serve speed across all matches played during the 2025 season. It serves as a proxy for overall serving power.

* $hgt$: Height (measured in **cm**). This represents a player's standing height in centimeters. Taller players benefit from a higher point of ball contact, a more favorable angle into the service box, and a longer swing arc, all of which are expected to generate greater serve speed. We hypothesize a positive relationship with serve speed.

* $age$: Age (measured in **years**). The player's age as of the 2025 season. Physical explosiveness and swing speed are generally expected to peak in a player's mid-to-late twenties and decline thereafter, leading us to expect a negative relationship between age and serve speed.

* $rank$: ATP ranking (**1–100**). This is the player's end-of-season ATP ranking. Given the development of tennis in the last years, we expect players with a higher ranked to have a faster serve, as part of a more complete game. However, because ranking may itself be considered a function of match performance, which is partly determined by serve speed,  including it might raise reverse causality concerns. 

* $frtserve$: First serve percentage (measured as a **%**). This value equals the proportion of first serves that land in the serving box. Players who serve closer to their physical limit, and at higher speeds, tend to miss first serves more frequently, so we expect a negative relationship between first serve percentage and average serve speed, because faster servers are more likely to miss their first serve.

* $dblfault$: Double faults (total **count**). This variable counts the total number of double faults committed during the season. Like first serve percentage, this variable functions as a behavioral proxy for serving aggression. Players who consistently serve at high speeds accept a greater risk of double faulting, so we expect a positive association between double faults and serve speed. It is not because faulting causes speed, but because both reflect the same underlying strategy.

* $rhsize$: Racket head size (measured in **cm²**). This variable measures the surface area of a player's racket head. We expected the racket head size and average speed to have a negative relationship, as players with a larger head size are expected to generate less power, also due to to an increased air resistance. However, we expect this variable not to be statistically significant, as professionals tend to standardize their equipment, meaning there is not much difference between the values for this variable in the sample.

* $rhand$: Handedness (1 = right-handed, 0 = left-handed). This is a dummy variable indicating whether a player serves with their right hand, or with their left hand. We expected this variable to matter for serve speed because tennis players that are left-handed usually prioritize lower speeds and more spin in their serves, to take advantage of their difference against right-handed players.

### Test Regression

```{r}
reg_test <- lm(avgspeed ~ age + hgt, data=tennis.data)
summary(reg_test)
```

  Before estimating the full model, we run preliminary regression on a convenience sample of the 21 ATP players for whom complete data was most readily available at the time. This pilot test regressed average serve speed on age and height only, and served two purposes: first, to verify that the data was structured and coded correctly before scaling to the full dataset; and second, to get an early indication on whether our two primary hypothesized predictors behaved as expected. Both age and height returned statistically significant coefficients with the anticipated signs: age was found to have a statistically significant negative effect on serve speed. Specifically, each additional year of age is associated with a decrease in average serve speed of 0.586 km/h, holding all other variables constant. This result is significant at the 95% confidence level, suggesting that as players grow older, their serve speed tends to decline — consistent with the expected reduction in physical explosiveness and swing speed that accompanies aging. 
Height was found to have a statistically significant positive effect on serve speed, and represents the strongest predictor in this initial model. Each additional centimeter of height is associated with an increase in average serve speed of 0.687 km/h, holding all other variables constant. This result is significant at the 99.9% confidence level, providing strong evidence that taller players are able to generate greater serve velocity. 
Both findings align with our initial expectations and lend early support to the hypothesis that physical attributes are primary determinants of serve speed among professional ATP players. However, with only 21 observations and 18 degrees of freedom, the sample is far too small to support reliable inference.

### Regressions 

$$ServeSpeed =\beta_0 + \beta_1 hgt + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 hgt + \beta_2 hgt^2 + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 age + \beta_2 hgt + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 age + \beta_2 hgt + \beta_3 (hgt)^2 + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 age + \beta_2 hgt + \beta_3 rhand + \beta_4 rhsize + \beta_5 fstserve + \beta_6 dblfault + \beta_7 small + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 rank + \beta_2 age + \beta_3 hgt + \beta_4 rhand + \beta_5 rhsize + \beta_6 fstserve + \beta_7 dblfault + \mu $$

```{r results='asis'}
heigth_reg <- lm(avgspeed ~ hgt, data = fulldata)

hgt_sq <- fulldata$hgt^2
heigthsq_reg <- lm(avgspeed ~ hgt + hgt_sq, data = fulldata)

hgtage_reg <- lm(avgspeed ~ age + hgt, data = fulldata)

hgtsqage_reg <- lm(avgspeed ~ age + hgt + hgt_sq, data = fulldata)

small <- ifelse(fulldata$avgspeed<205, 1, 0)
outlier_reg <- lm(avgspeed ~ age + hgt + rhand + rhsize + frtserve + dblfault + small, data = fulldata)

final_reg <- lm(avgspeed ~ rank + age + hgt + rhand + rhsize + frtserve + dblfault, data = fulldata)

htmlreg(list(heigth_reg, heigthsq_reg, hgtage_reg, hgtsqage_reg, outlier_reg, final_reg),
        stars=c(0.01,0.05,0.1),
        caption = "Dep Var: R&D Intensity",
        star.symbol = "\\*",
        caption.above = TRUE, 
        custom.model.names=c("Height", "Height^2", "Height&Age", "Height^2&Age", "No Outliers", "Final"),
        digits=4)
```


### Regression Results

We discuss the findings of the table with all regressions variable by variable, tracing how each coefficient evolves as the model is built out.

* $ranking (rank)$: Ranking is positive (0.035) but not statistically significant (p = 0.137). Higher-ranked players do not appear to serve meaningfully faster than lower-ranked ones within the top 100, once physical and behavioral characteristics are accounted for. This is consistent with the broader finding that serve speed is a specific, height-driven attribute rather than a general marker of overall tennis quality — players can reach and sustain a top-100 ranking through many different combinations of skills, of which serve speed is only one.

* $Age (age)$: Age presents one of the more instructive contrasts in this analysis. In the preliminary test regression on our 21-player convenience sample, age appeared significant at the 95% level with a coefficient of −0.586, suggesting that each additional year of age reduced serve speed by over half a kilometer per hour. This seemed to confirm our hypothesis that physical decline meaningfully affects serving power.
However, once the full 100-player dataset is used, this result disappears entirely. Across every full-sample specification — whether paired only with height, or alongside the complete set of variables — age carries a negative coefficient in the range of −0.103 to −0.128, but never approaches statistical significance (p-values ranging from 0.341 to 0.433). The contrast between the test regression and the full-sample results is a deliberate and instructive reminder of the risks of small-sample inference: a finding that appeared robust at n=21 was simply an artifact of limited data.
The most likely explanation for age's insignificance in the full sample is a selection effect. The ATP top 100 is not a random cross-section of players at various stages of physical decline — it is, by definition, an elite group. Older players who remain ranked in the top 100 at age 33, 35, or beyond have done so precisely because they have preserved their physical capabilities. Players who have genuinely declined in serve speed are no longer in this sample. Within this compressed, self-selected group, age loses its predictive power.

* $Height (hgt)$: Height is the central finding of this paper, and its story across all specifications is remarkably consistent. In the simplest model — regressing serve speed on height alone — each additional centimeter of height is associated with a 0.766 km/h increase in average serve speed, significant at the 99.9% confidence level. This single variable explains nearly 50% of the total variation in serve speed (R² = 0.495), an exceptional result for a one-variable regression that immediately signals height as the dominant structural driver of serving power.
To test whether this relationship exhibits diminishing returns — that is, whether extreme height eventually stops paying off in serve speed — we introduce a quadratic term (hgt²). It is not statistically significant (p = 0.942), and the adjusted R² slightly decreases, penalizing the added complexity. The relationship between height and serve speed is linear throughout the range observed in our sample: every centimeter yields roughly the same return, with no evidence that very tall players see a smaller marginal benefit. This is the reason why the squared value of height does not appear in the final regression.
As additional variables are introduced, height's coefficient remains virtually unchanged — 0.776 when age is added, 0.781 in the full model with all variables included — and it retains 99.9% significance in every single specification. This stability is not a minor technical detail; it is strong evidence that height's effect is not being confounded by any of the other variables in our dataset. For a coach or scout, the practical implication is clear: the difference between a 180 cm and a 198 cm player carries a predicted serve speed advantage of approximately 13.8 km/h — a gap that is largely structural, persistent, and not easily compensated for through training or equipment.

* $Hand (rhand)$: Right-handed players are estimated to serve approximately 2.1–2.2 km/h faster than left-handed players across the specifications that include this variable, but the coefficient never reaches statistical significance (p-values around 0.165–0.187). The sample composition largely explains this: with 87 right-handed players and only 13 left-handed players, the test simply lacks the statistical power to detect even a moderate true effect. A meaningful test of handedness as a determinant of serve speed would require a more balanced sample, potentially drawing from a broader population of players beyond the top 100.

* $Racket Head Size (rhsize)$: Racket head size is positive across all specifications where it is included, with a coefficient ranging from 0.069 to 0.109. In the final model it reaches marginal significance at the 10% level (p = 0.088), though it does not clear the 5% threshold in any specification. The direction is somewhat counterintuitive relative to our original hypothesis — we expected a negative relationship based on air resistance concerns — but a positive sign is plausible if a larger sweet spot allows players to make more consistent high-speed contact, converting aggressive swing speeds into faster ball velocity more reliably.
That said, the result is not robust enough to draw firm conclusions. The range of racket head sizes in the sample is narrow (600–645 cm²), which limits the model's ability to detect this effect precisely. This variable warrants further investigation in a dataset with greater equipment variation.

* $First Serve % (frtserve)$: First serve percentage carries a negative coefficient in every specification (ranging from −0.174 to −0.231), consistent with the intuition that players who serve closer to their physical limit will miss more first serves. However, it never reaches statistical significance (p-values around 0.145–0.265). The likely explanation is that this variable and double faults are capturing the same underlying construct — serve aggression — and in the presence of both, neither fully dominates. The double fault variable appears to carry this signal more cleanly, leaving first serve percentage without independent explanatory power.

* $Double Faults (dblfault)$: Double faults emerge as the only variable besides height to reach conventional statistical significance in the full model. Each additional double fault over the course of a season is associated with a 0.029 km/h increase in average serve speed (p = 0.012, significant at the 95% level). In the no-outlier specification, the coefficient is slightly smaller at 0.017 but remains significant at the 10% level.
The interpretation here is behavioral, not causal. Double faults do not make players serve faster — rather, players who choose to serve at higher speeds on both first and second deliveries naturally commit more double faults as a consequence of that aggression. The double fault count functions as a proxy for serving risk tolerance: it captures the strategic decision to prioritize speed over consistency. A coach should read this not as a reason to accept more double faults, but as confirmation that elite servers are consciously operating at the edge of their delivery margin.

* Outlier Check: Using a boxplot of average serve speeds, we identified a small number of players at the upper end of the distribution — those averaging above 205 km/h — as potential outliers that could be exerting disproportionate influence on the results. To test this, we construct a dummy variable equal to one for players below this threshold (capturing the non-outlier group) and include it as an additional control.
The dummy itself is negative and significant at the 10% level (−7.672, p = 0.066), confirming that the flagged observations do serve measurably faster than the rest of the sample. Controlling for the outliers in the regression does not impact the significance of the coefficients. Crucially, none of the other coefficients change meaningfully in sign, magnitude, or significance. Height retains its 99.9% significance, double faults remain significant, and all other variables remain non-significant. The core findings of this paper are not an artifact of a handful of extreme servers — they hold cleanly across the full distribution of the sample.

### How did we find outliers?

```{r}
boxplot(fulldata$avgspeed,
        data = fulldata,
        main="Sample Outliers",
        xlab = "",
        ylab = "avgspeed",
        col = "orange",
        border = "brown"
        )

small <- ifelse(fulldata$avgspeed<205, 1, 0)
```


### Ultimate Diagnosis

```{r}
#BP and White Test for age and height
bptest(avgspeed ~ age + hgt, data = fulldata, studentize=TRUE)

reg_3 <- lm(avgspeed ~ age + hgt, data = fulldata)
reg3_fit <- fitted(reg_3)
reg3_res <- residuals(reg_3)
reg3_res_sq <- reg3_res^2
reg3_fit_sq <- reg3_fit^2

reg3_w_test <- lm(reg3_res_sq ~ reg3_fit + reg3_fit_sq, data = fulldata)

reg3_n <- nrow(fulldata)
reg3_r2 <- summary(reg3_w_test)$r.squared
reg3_lm_stat <- reg3_n*reg3_r2
reg3_lm_stat

pchisq(reg3_lm_stat, df=2, lower.tail = FALSE)

plot(fitted(reg_3), residuals(reg_3))

# BP and White Test for Final Regression
bptest(avgspeed ~ age + hgt + hgt_sq + rhand + rhsize + frtserve + dblfault, data = fulldata, studentize=TRUE)

reg_6 <- lm(avgspeed ~ rank + age + hgt + rhand + rhsize + frtserve + dblfault, data = fulldata)
reg6_fit <- fitted(reg_6)
reg6_res <- residuals(reg_6)
reg6_res_sq <- reg6_res^2
reg6_fit_sq <- reg6_fit^2

reg6_w_test <- lm(reg6_res_sq ~ reg6_fit + reg6_fit_sq, data = fulldata)

reg6_n <- nrow(fulldata)
reg6_r2 <- summary(reg6_w_test)$r.squared
reg6_lm_stat <- reg6_n*reg6_r2
reg6_lm_stat

pchisq(reg6_lm_stat, df=2, lower.tail = FALSE)

plot(fitted(reg_6), residuals(reg_6))
```

A key assumption of OLS regression is homoskedasticity — that the variance of the error term is constant across all observations. If this assumption is violated, standard errors become unreliable, and the significance levels we report for our coefficients can no longer be trusted. We test for heteroskedasticity formally across our two primary specifications using two complementary procedures: the Breusch-Pagan (BP) test and the White test.

The Breusch-Pagan test examines whether the squared residuals of the model can be explained by the independent variables. A significant result indicates that error variance is systematically related to the regressors — a sign of heteroskedasticity. The White test is a more general, flexible version of this procedure that also allows for non-linear and interaction effects in the error variance, making it less sensitive to specific assumptions about the form of heteroskedasticity.

For the Height and Age model (Regression 3): the Breusch-Pagan test returns a statistic of BP = 2.083 with 2 degrees of freedom and a p-value of 0.353. The White test returns an LM statistic of 0.609 with a corresponding p-value of 0.738. Both p-values are far above conventional significance thresholds. We fail to reject the null hypothesis of homoskedasticity under both tests, meaning the residuals from this model display constant variance across all fitted values.

For the Final Model (Regression 4): the Breusch-Pagan test returns BP = 7.893 with 7 degrees of freedom and a p-value of 0.342. The White test returns an LM statistic of 0.501 with a p-value of 0.779. Again, both tests fail to reject the null hypothesis of constant variance. The residual-versus-fitted plots for both models further confirm this: the scatter of residuals around zero is roughly uniform across the range of predicted values, with no discernible fan shape, curvature, or systematic pattern that would indicate non-constant variance.

The implication is straightforward and important: both models are homoskedastic. The standard errors reported in all regression tables are valid, and the significance levels attributed to each coefficient can be trusted without correction. No robust standard error adjustment (such as HC standard errors) is necessary, and the inference drawn throughout this paper rests on a statistically sound foundation.

The primary endogeneity concern in this model is the omitted variable of toss height/point of contact, which we already identified in our original variable list. This variable is positively correlated with both height (taller players make contact higher with the ball) and serve speed, meaning height's coefficient may carry a slight upward bias. Since toss height is unavailable to us from the official ATP data, it cannot be directly controlled for in our regression. However, because it is largely a function of height itself, we believe the bias to be insignificant. A second concern is reverse causality between double faults and serve speed, as players usually decrease serve speed because they are already double-faulting frequently. It becomes a matter of strategy more than a matter of physical capabilities. 

## Conclusion

This paper examined the determinants of average serve speed among the top 100 ATP players using OLS regression analysis. Across all model specifications, height emerged as the singular robust predictor of serve speed, with each centimeter of additional height associated with approximately 0.76 km/h of additional speed. Double faults — interpreted here as a proxy for serve aggression — were the only other variable to reach statistical significance. Age, our dummy variable for being right or left handed, racket head size, and first-serve percentage were not statistically significant once height was included in the model.
These results carry meaningful implications. For sports analysts and bettors, serve speed is largely a function of a fixed physical trait, making it a predictable input for match forecasting. For equipment manufacturers, the lack of significance for racket head size suggests that at the professional level, players have already optimized their gear, and performance differences are driven by the athlete, not the equipment.
The primary limitation of this study is sample size (n=100) and the cross-sectional nature of the data. Longitudinal data tracking individual players over time would allow for stronger causal claims, particularly regarding age-related decline. Future work should also explore the inclusion of toss height or point-of-contact data, which remains the key unobservable in this analysis.

