  ---
title: "A Study on the Average Serve Speed of Top Tennis Players, and the Factors that Affect them"
author: "GJornert, EMelindo, HArnold"
date: "2026-02-02"
output: html_document
---

```{r setup, include=FALSE}

library(tidyverse)
library(wooldridge)
library(broom)
library(texreg)
library(readxl)
library(xtable)
library(dplyr)
library(corrr)

library(lmtest)
library(sandwich)
library(hettest)
library(vars)

library(nlme)

knitr::opts_chunk$set(echo = TRUE)

```

### The Determinants of Serve Speed in Professional Tennis 

In tennis, the serve is far more than just the starting point of a rally — it is one of the most decisive shots in the game. A powerful serve can earn a player a free point, limit their opponent's ability to return aggressively, and dictate the flow of the entire match. At the highest levels of professional tennis, serve speed is considered a critical competitive advantage, which raises an important question: what factors actually determine how fast an ATP player can serve?

This paper investigates the determinants of average serve speed among professional ATP players through a series of regression analyses. We examine both physical characteristics — including height, weight, and age — as well as experiential and equipment-related variables, such as years on tour and racket specifications. Our central research question is: to what extent do a player's physical attributes, professional experience, and equipment explain variation in average serve speed? More specifically, we ask whether natural physical factors or acquired, controllable factors are the stronger predictors of performance.

Understanding what drives serve speed also carries meaningful economic implications. If serve speed is primarily determined by physical traits, it becomes a more predictable variable for sports analysts, bettors, and scouts — offering an edge in forecasting match outcomes based on player profiles. On the other hand, if equipment plays a significant role, racket manufacturers have a tangible, data-backed basis for marketing performance claims, much as footwear and apparel brands have done in other sports. Either finding has real consequences for how money flows through the tennis industry.

We hypothesize that physical attributes — particularly height — will be the strongest predictor of serve speed, as taller players benefit from a more favorable serving angle and a longer swing arc, allowing them to generate greater velocity more naturally. Weight and age are also expected to be significant physical factors, with serve speed likely peaking during a player's prime years and declining thereafter. While we expect racket specifications — such as racket weight and string tension — to have a meaningful impact on serve speed as well, we anticipate their effect to be secondary to that of physical traits, since equipment can optimize performance but cannot fully compensate for the bio mechanical advantages conferred by a player's body. To test these hypotheses, our data set includes player-level statistics from the ATP tour, comprising physical measurements, equipment specifications, and career statistics.
Our findings confirm that height is the dominant and most robust determinant of serve speed, while most other variables — including age, equipment, and handedness — do not reach statistical significance in the full sample.

### Liturare Review

Unsupreiseianly other reseashess in the field of sports have been increastesd in the how the serve speed is effected by the player.

#### Tennis Serve Biomechanics

Tennis Serve Biomechanics, Joint Load Mechanics, and Overuse Injuries: A Narrative Review by Caroline Martin, Pierre Touzard, Loic Fourel, and Babette M. Pluim examines how the body is affected by the act of serving. finding that improper form and technique induce eminently traumatic mechanical joint loading, which can lead to injury over time. To mitigate joint damage, the paper finds that teaching proper serving technique, choosing appropriate equipment, and training conditions can reduce the effects of a tennis serve on the body. providing a reason behind the link between a player's serve, their equipment, and their body.

#### The Role of Toss Zenith and Impact Height in the Serve Speed in Junior Tennis Players.

The Role of Toss Zenith and Impact Height in the Relationship Between Shoulder Rotation Strength and Serve Speed in Junior Tennis Players, by Jan Vacek, Michal Vagner, Jan Malecek, Jachym Simsa, Vladan Olah, and Petr Stastny. The paper details how various factors, such as zenith height, impact height, and the zenith–impact differential of an opening, affect Serve Speed in Junior Tennis Players. Since this research is based on junior players, the findings are likely to be similar to those of professional players. Their findings show that the height of the Impact showed a positive association with serve speed, as well as with the execution of the opening toss and its consistency, significantly increasing serve speed.

#### Age and sex-related upper body performance in young tennis players.

Age- and sex-related upper-body performance differences in competitive young tennis players, by Jaime Fernandez-Fernandez, Fabio Yuzo Nakamura, Victor Moreno-Perez, Alejandro Lopez-Valenciano, Juan Del Coso, and Cesar Gallo-Salazar. The study aimed to analyze the physical abilities of young male and female tennis players to identify relationships among physical variables and serve speed. The study directly uses serve speed as a factor, making it a relevant reference for comparing with our own findings. The study found that service speed increased across all players, regardless of sex.

### Data and variables

How do physical characteristics, equipment, ATP/WTA tour activity, and ranking affect the average serve speed of a tennis professional? 
Average serve speed in top 100 -

* hgt - height - positive correlation
* wgt - weight - positive correlation (multicollinearity w height)
* age - age - negative
* rank - ATP/WTA ranking - positive, 
* mtch - number of matches played/year - unknown 
* rhsize - racket head size - negative, 
* fstserve - first serve percentage - negative 
* dblfault - double faults - positive
* gen - gender - male (1) players serve faster than female (0) player (standby for now).
* rhand - Right(1)/left(0)-handed - unknown
* Unobservable (omitted) variable:*
* Toss height (point of impact) - positive correlation with height and serve speed (positive bias).

```{r} 
# Partial data
tennis.data <- read_excel("econ-research_test.xlsx")

summary(tennis.data)
```
```{r}
#rename column by name
colnames(tennis.data)[colnames(tennis.data) == '1serve'] <- 'frtserve'
colnames(tennis.data)[colnames(tennis.data) == '2fault'] <- 'dblfault'

```

```{r}
# Completed data set
fulldata <- read_excel("econ.research_v1.xlsx")

summary(fulldata)
```
```{r}
#rename column by name
colnames(fulldata)[colnames(fulldata) == '1serve'] <- 'frtserve'
colnames(fulldata)[colnames(fulldata) == '2faults'] <- 'dblfault'

```

### List of Future Regressions 

$$ServeSpeed =\beta_0 + \beta_1 hgt + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 hgt + \beta_2 hgt^2 + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 age + \beta_2 hgt + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 age + \beta_2 hgt + \beta_3 (hgt)^2 + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 rank + \beta_2 age + \beta_3 hgt + \beta_4 rhand + \beta_5 mtch + \beta_6 rhsize + \beta_7 fstserve + \beta_8 dblfault + \mu $$

### Test Regression

```{r}
reg_test <- lm(avgspeed ~ age + hgt, data=tennis.data)
summary(reg_test)
```

* As an initial test, a regression was run on a sample of the first 20 ATP players for whom data was most readily available, incorporating age and height as independent variables. The results are as follows:

Age was found to have a statistically significant negative effect on serve speed. Specifically, each additional year of age is associated with a decrease in average serve speed of 0.586 km/h, holding all other variables constant. This result is significant at the 95% confidence level, suggesting that as players grow older, their serve speed tends to decline — consistent with the expected reduction in physical explosiveness and swing speed that accompanies aging.

Height was found to have a statistically significant positive effect on serve speed, and represents the strongest predictor in this initial model. Each additional centimeter of height is associated with an increase in average serve speed of 0.687 km/h, holding all other variables constant. This result is significant at the 99.9% confidence level, providing strong evidence that taller players are able to generate greater serve velocity. This is consistent with our hypothesis: greater height raises the point of ball contact, creates a more favorable angle into the service box, and allows for a longer swing arc — all of which contribute to higher serve speeds.

Both findings align with our initial expectations and lend early support to the hypothesis that physical attributes are primary determinants of serve speed among professional ATP players. It should be noted, however, that this regression is based on a limited sample of 20 observations, and the results should be interpreted with caution until the full data set is incorporated.

### Regression 1 & 2: Height

$$ServeSpeed =\beta_0 + \beta_1 hgt + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 hgt + \beta_2 hgt^2 + \mu $$


```{r}
heigth_reg <- lm(avgspeed ~ hgt, data = fulldata)
summary(heigth_reg)

## Height Squared

hgt_sq <- fulldata$hgt^2
heigthsq_reg <- lm(avgspeed ~ hgt + hgt_sq, data = fulldata)
summary(heigthsq_reg)

## heteroskedasticity test
bptest(avgspeed ~ hgt + hgt_sq, data = fulldata, studentize = TRUE)
```

The squared term for height was tested but found to be insignificant, confirming a linear relationship between height and serve speed. For this reason, it will be excluded from the future regressions.

### Regression 3: Height and Age

$$ServeSpeed =\beta_0 + \beta_1 age + \beta_2 hgt + \mu $$

$$ServeSpeed =\beta_0 + \beta_1 age + \beta_2 hgt + \beta_3 (hgt)^2 + \mu $$

```{r}
hgtage_reg <- lm(avgspeed ~ age + hgt, data = fulldata)
summary(hgtage_reg)

hgtsqage_reg <- lm(avgspeed ~ age + hgt + hgt_sq, data = fulldata)
summary(hgtsqage_reg)

```

Checking for heteroskedasticity: 

```{r}

reg_3 <- lm(avgspeed ~ age + hgt, data = fulldata)
summary(reg_3)

bptest(avgspeed ~ age + hgt, data = fulldata, studentize=TRUE)

reg3_fit <- fitted(reg_3)
reg3_res <- residuals(reg_3)
reg3_res_sq <- reg3_res^2
reg3_fit_sq <- reg3_fit^2

reg3_w_test <- lm(reg3_res_sq ~ reg3_fit + reg3_fit_sq, data = fulldata)


reg3_n <- nrow(fulldata)
reg3_r2 <- summary(reg3_w_test)$r.squared
reg3_lm_stat <- reg3_n*reg3_r2
reg3_lm_stat

pchisq(reg3_lm_stat, df=2, lower.tail = FALSE)

plot(fitted(reg_3), residuals(reg_3))
```

The results from both BP test and White test prove that this model is homoskedastic. Both tests confirm the model's residuals have constant variance. Standard errors and significance levels reported in the regression output are therefore reliable and do not require correction.

### Regression 4: All Variables

```{r}
every_reg <- lm(avgspeed ~ rank + age + hgt + rhand + rhsize + frtserve + dblfault, data = fulldata)
summary(every_reg)
```

* Height (hgt): Statistically significant at the 99.9% level (p < 0.001). Each additional centimeter of height increases average serve speed by 0.761 km/h, holding all else constant. This is the dominant predictor in the model and is consistent across every specification tested.
* Double Faults (dblfault): Significant at the 95% level (p = 0.042). Each additional double fault is associated with a 0.019 km/h increase in serve speed. This likely reflects a behavioral pattern — players who serve harder accept a higher fault risk, so double faults are a proxy for serve aggression rather than a cause of speed.
* Age (age): Negative coefficient (−0.103) but not statistically significant (p = 0.433). Age does not independently explain serve speed variation once height is controlled for. This contradicts our initial test-sample result and suggests the earlier significance was driven by the small n=21 sample.
* Dummy Variable (rhand): Positive coefficient (2.209) suggesting right-handers serve ~2.2 km/h faster, but not significant (p = 0.165). Sample is heavily skewed right-handed (87%), limiting statistical power.
* Racket Head Size (rhsize): Positive (0.092) but not significant (p = 0.144). Equipment does not appear to meaningfully explain serve speed differences at the professional level, consistent with the idea that all top players optimize their equipment.
* First Serve % (frtserve): Negative (−0.224) but not significant (p = 0.145). The direction is intuitive — players who serve harder make a lower percentage — but the effect is absorbed by other variables.
The overall fit of the model can be seen through the adjusted R^2 = 0.517, meaning the model explains roughly 52% of the variation in serve speed. 

Testing for heteroskedasticity

```{r}

reg_6 <- lm(avgspeed ~ age + hgt + rhand + rhsize + frtserve + dblfault, data = fulldata)
summary(reg_6)

bptest(avgspeed ~ age + hgt + hgt_sq + rhand + rhsize + frtserve + dblfault, data = fulldata, studentize=TRUE)

reg6_fit <- fitted(reg_6)
reg6_res <- residuals(reg_6)
reg6_res_sq <- reg6_res^2
reg6_fit_sq <- reg6_fit^2

reg6_w_test <- lm(reg6_res_sq ~ reg6_fit + reg6_fit_sq, data = fulldata)


reg6_n <- nrow(fulldata)
reg6_r2 <- summary(reg6_w_test)$r.squared
reg6_lm_stat <- reg6_n*reg6_r2
reg6_lm_stat

pchisq(reg6_lm_stat, df=2, lower.tail = FALSE)

plot(fitted(reg_6), residuals(reg_6))
```

This model does not suffer from heteroskedasticity. We can trust the significance/standard error to be correct.

The primary endogeneity concern in this model is the omitted variable of toss height/point of contact, which we already identified in our original variable list. This variable is positively correlated with both height (taller players make contact higher with the ball) and serve speed, meaning height's coefficient may carry a slight upward bias. Since toss height is unavailable to us from the official ATP data, it cannot be directly controlled for in our regression. However, because it is largely a function of height itself, we believe the bias to be insignificant. A second concern is reverse causality between double faults and serve speed, as players usually decrease serve speed because they are already double-faulting frequently. It becomes a matter of strategy more than a matter of physical capabilities. 

```{r}
boxplot(fulldata$avgspeed,
        data = fulldata,
        main="Average Serve Speed",
        xlab = "",
        ylab = "Height",
        col = "orange",
        border = "brown"
        )
```

```{r results='asis'}
small <- ifelse(fulldata$hgt<205, 1, 0)

reg_6 <- lm(avgspeed ~ age + hgt + rhand + rhsize + frtserve + dblfault, data = fulldata)
summary(reg_6)
reg_7 <- lm(avgspeed ~ age + hgt + rhand + rhsize + frtserve + dblfault + small, data = fulldata)
summary(reg_7)

htmlreg(list(reg_6, reg_7),
        stars=c(0.01,0.05,0.1),
        caption = "Dep Var: R&D Intensity",
        star.symbol = "\\*",
        caption.above = TRUE, 
        custom.model.names=c("Baseline", "No Outlier"),
        digits=4)
```

Taking out the outliers does not impact the significance of the coefficients. 

## Conclusion

This paper examined the determinants of average serve speed among the top 100 ATP players using OLS regression analysis. Across all model specifications, height emerged as the singular robust predictor of serve speed, with each centimeter of additional height associated with approximately 0.76 km/h of additional speed. Double faults — interpreted here as a proxy for serve aggression — were the only other variable to reach statistical significance. Age, our dummy variable for being right or left handed, racket head size, and first-serve percentage were not statistically significant once height was included in the model.
These results carry meaningful implications. For sports analysts and bettors, serve speed is largely a function of a fixed physical trait, making it a predictable input for match forecasting. For equipment manufacturers, the lack of significance for racket head size suggests that at the professional level, players have already optimized their gear, and performance differences are driven by the athlete, not the equipment.
The primary limitation of this study is sample size (n=100) and the cross-sectional nature of the data. Longitudinal data tracking individual players over time would allow for stronger causal claims, particularly regarding age-related decline. Future work should also explore the inclusion of toss height or point-of-contact data, which remains the key unobservable in this analysis.

## Resouces

* Martin, Caroline, et al. “Tennis serve biomechanics, Joint Load Mechanics and overuse injuries: A narrative review.” Sports Medicine - Open, vol. 12, no. 1, Dec. 2026, p. NA. Gale Academic OneFile, http://dx.doi.org/10.1186/s40798-026-01015-y. 

* Vacek, Jan, et al. "The Role of Toss Zenith and Impact Height in the Relationship Between Shoulder Rotation Strength and Serve Speed in Junior Tennis Players." Journal of Functional Morphology and Kinesiology, vol. 10, no. 4, Dec. 2025. Gale Academic OneFile, dx.doi.org/10.3390/jfmk10040438. Accessed 24 Apr. 2026.
